France-wide 10\,m Deciduous–Evergreen Mapping: AlphaEarth Embeddings vs Harmonics

### 2025-10-12 article/article.tex
add new article version

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Frontiers LaTeX template – v2 (2025-04-10)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[utf8]{FrontiersinHarvard}
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage[onehalfspacing]{setspace}
\usepackage{float}
\usepackage{multirow}
\linenumbers

\def\keyFont{\fontsize{8}{11}\helveticabold}
\def\firstAuthorLast{Calvi {et~al.}}
\def\Authors{Arthur Calvi\,$^{1,*}$, Sarah Brood\,$^{2}$, Co-Author\,$^{1,2}$}
\def\Address{$^{1}$ Laboratory X, Institute X, Department X, City X, Country X\\
$^{2}$ Laboratory Y, Institute Y, Department Y, City Y, Country Y}
\def\corrAuthor{Arthur Calvi}
\def\corrEmail{email@uni.edu}

@@

\title[France tree-phenology map]{France-wide 10\,m Deciduous-Evergreen Map from Sentinel-2 Harmonic Features}

@@

\begin{abstract}
Effective forest monitoring requires distinguishing normal seasonal changes from disturbance events. Löw et al. (2020) showed that incorporating phenological information—whether trees are deciduous or evergreen—significantly reduces false disturbance alerts in satellite-based monitoring systems. Europe lacks this fundamental layer: while Copernicus DLT classifies broadleaf versus conifer at 10\,m resolution, it cannot distinguish evergreen oak from deciduous beech. France's BD Forêt V2 provides phenological classes but remains static (2005–2019 compilation). We present France's first annual 10\,m deciduous-evergreen map using harmonic analysis of Sentinel-2 time series. Our method compresses each pixel's annual cycle into 14 harmonic features capturing seasonal amplitude, timing, and consistency. A Random Forest classifier with 50 trees and eco-region balanced cross-validation achieves 90\% accuracy on 14 million reference pixels. The approach produces national maps in under an hour, enabling annual updates that existing products cannot provide.

\keyFont{\section{Keywords:} deciduous-evergreen classification, harmonic analysis, Sentinel-2, Random Forest, France}
\end{abstract}

@@

Satellite-based forest monitoring systems face a fundamental challenge: distinguishing genuine disturbance events from normal seasonal changes. Löw et al. (2020) demonstrated that incorporating phenological information—whether trees are deciduous or evergreen—eliminated most false disturbance alerts in Austrian forests. Similarly, Hargrove et al. (2009) showed that the U.S. eMODIS monitoring system required phenology filters to separate true forest threats from seasonal transitions. These studies highlight a critical need: operational forest surveillance requires knowing not just where trees grow, but how they behave through the seasons.

Yet Europe lacks this fundamental layer. The Copernicus Dominant Leaf Type product (EU, 2024a) tells us whether a pixel contains broadleaf or coniferous trees at 10\,m resolution, but a Mediterranean cork oak (evergreen broadleaf) appears identical to a northern beech (deciduous broadleaf). France's BD Forêt V2 (IGN, 2024) does classify deciduous versus evergreen, but as a static compilation from 2005–2019 aerial photos, it cannot track annual variations or recent disturbances.

This distinction matters because deciduous and evergreen forests represent fundamentally different ecological strategies. Deciduous species maximize photosynthetic capacity during favorable seasons but sacrifice year-round productivity, while evergreen species maintain steady-state photosynthesis across seasons with lower peak rates (Jarvis \& Leverenz, 1983). These physiological differences create distinct temporal signatures in satellite observations, making phenological classification both scientifically meaningful and practically achievable. Moreover, temperate forests increasingly form complex deciduous-evergreen mosaics shaped by disturbance regimes and species interactions (Frelich, 2002), requiring high-resolution mapping to capture fine-scale heterogeneity essential for forest management and ecological understanding.

This paper presents France's first annual 10\,m deciduous-evergreen map, built on three key innovations:
\begin{enumerate}
    \item \textbf{Harmonic compression} — We distill each pixel's annual time-series into 14 interpretable features capturing phenological patterns
    \item \textbf{Eco-region balanced training} — Our cross-validation ensures model generalization across France's diverse ecological gradients
    \item \textbf{Operational efficiency} — The entire country processes in under an hour, enabling annual updates
\end{enumerate}

Our method achieves 90\% accuracy while remaining computationally efficient and scientifically interpretable. This work builds on the demonstrated feasibility of national-scale automated mapping using Sentinel-2 time-series and Random Forest classifiers, as shown by Inglada et al. (2017) who achieved 86\% accuracy mapping 17 land cover classes across France, establishing the foundation for operational satellite-based monitoring at country scale.

\section{Related Work}

Time-series analysis of vegetation has evolved along two paths: precise but computationally intensive curve-fitting methods, and efficient but less interpretable machine learning approaches. We chose a middle way that balances accuracy with operational feasibility.

Traditional phenology extraction tools like TIMESAT and HR-VPP excel at pinpointing exact green-up and senescence dates through logistic curve fitting (Jönsson \& Eklundh, 2002; EU, 2024b). However, their pixel-by-pixel optimization becomes prohibitively slow at 10\,m resolution across national scales. Change detection algorithms like BEAST and CCDC can identify abrupt forest disturbances but require GPU clusters and provide more detail than needed for binary deciduous-evergreen classification.

At the other extreme, deep learning methods—whether convolutional networks or LSTMs processing raw image stacks—can leverage spatial context and learn complex patterns. Yet they demand massive labeled datasets (typically >100 million pixels) and operate as black boxes, making validation and interpretation challenging for operational forestry.

Harmonic regression offers an elegant compromise. By decomposing time-series into sinusoidal components, we capture the essence of phenological cycles in a handful of interpretable features: how strongly vegetation oscillates (amplitude), when it peaks (phase), and how consistently it follows the pattern (residuals). This Fourier approach has proven effective in crop mapping and land cover classification, with recent advances demonstrating its potential for forest species mapping. Francini et al. (2024) successfully combined Sentinel-2 harmonic analysis with national forest inventory data to map tree species distributions, establishing that integrating satellite phenology metrics with ground observations represents a credible, state-of-the-art approach for operational forest monitoring.

Recent work has specifically demonstrated the power of phenology-based approaches for forest type classification. Li et al. (2023) achieved accurate evergreen-deciduous separation at 10\,m resolution across China using a phenology index based on seasonal NDVI differences, while Bolton et al. (2020) developed continental-scale land surface phenology algorithms combining Landsat-8 and Sentinel-2, demonstrating fine-scale vegetation phenology mapping with particular success in capturing deciduous forest seasonality. These studies establish that leveraging seasonal differences in vegetation indices provides a robust foundation for distinguishing forest leaf types at large scales.

Our contribution demonstrates that just 14 carefully selected harmonic features can distinguish deciduous from evergreen forests as accurately as methods requiring orders of magnitude more computation or training data, while providing the annual updates that existing static products cannot deliver.

@@

\subsection{Study Area}
We mapped metropolitan France across its full ecological gradient—from Atlantic coastal forests to Alpine timber lines, Mediterranean maquis to continental oak stands. The country's 11 eco-regions (Jarvis et al., 1983) provided our spatial framework, ensuring our model learned phenological patterns from every forest type (Figure \ref{fig:training_tiles}).

% Note: eco_region_map.png not available - using tile visualization instead
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{France's training tiles distributed across eleven eco-regions, ensuring representation of diverse forest phenology patterns from oceanic to mediterranean climates.}
    \label{fig:training_tiles}
\end{figure}

\subsection{Satellite Observations}
We assembled monthly cloud-free mosaics from the Sentinel-2 archive (2017–2024), selecting images with less than 10\% cloud cover. Beyond the standard spectral bands (blue through shortwave infrared), we computed four vegetation indices known to capture phenological signals: NDVI for overall greenness, EVI for enhanced sensitivity in dense canopies (Huete et al., 2002), NBR for moisture content (Roy et al., 2006), and CRSWIR (Continuum Removal Short-Wave Infrared) for vegetation water stress detection.

The CRSWIR index applies continuum removal technique to SWIR bands, maximizing spectral contrast associated with absorption features by normalizing reflectance values relative to a convex hull envelope calculated from neighboring spectral bands (Dutrieux et al., 2021). This index strongly correlates with vegetation water content, with low CRSWIR values indicating high vegetation water content and increasing CRSWIR values signaling early physiological stress.

\subsection{Building a Balanced Reference Dataset}

Creating training data for 14 million pixels required careful orchestration of multiple sources while maintaining geographic and ecological balance.

We assembled training data from multiple sources: high-quality in-situ polygons from PureForest inventory, RENECOFOR permanent plots, and IGN's LiDAR-validated stands, supplemented by BD Forêt V2 polygons after strict quality control (50\,m buffer from in-situ data, boundary shrinking to avoid edge effects).

Our sampling strategy partitioned France into 2.5\,km × 2.5\,km tiles, selecting those with minimum 10 reference pixels while maintaining 5\,km separation to reduce spatial autocorrelation. A weighted sampling approach balanced representation across eco-regions, with weights calculated as the ratio of effective forest area fraction to dataset fraction, ensuring Mediterranean and Alpine forests contributed proportionally despite smaller absolute coverage.

The final dataset of 14.1 million pixels achieved remarkable diversity: 15 genera and 30 species, with oak (33\%) and pine (12\%) most common but no other genus exceeding 7\%. While deciduous pixels outnumber evergreen (75\% to 25\%), we addressed this imbalance through eco-region weighted sampling and balanced class weights during training.

\subsection{The Power of Harmonic Analysis}

We transformed each pixel's annual cycle into harmonic components using Fourier analysis, following the established framework for extracting seasonal vegetation dynamics from satellite time-series (Jönsson \& Eklundh, 2002). For every vegetation index, we fit up to two harmonic waves:

$$ f(t) = \sum_{i=1}^{2} \left[ A_i \cos(2\pi i t) + B_i \sin(2\pi i t) \right] + C $$

Cloud-contaminated observations receive lower weights based on Sentinel-2's cloud probability. The fitting process yields interpretable features: amplitude (seasonal strength), phase (timing encoded as cos/sin components), offset (baseline level), and residual variance (model consistency). This approach ensures that key seasonal differences between species—such as the timing of peak greenness and the magnitude of seasonal variation—are systematically quantified rather than derived ad hoc.

Circular encoding of phase prevents artificial splitting of temporally adjacent phenologies (e.g., December-January), ensuring Random Forest models respect seasonal continuity.

\subsection{Feature Selection and Model Training}

From 32 possible harmonic features (4 indices × 8 metrics including circular phase encoding), recursive feature elimination with cross-validation identified the 14 most informative:

\begin{itemize}
    \item \textbf{NDVI}: First harmonic amplitude, phase (cos/sin), offset, and residual variance (5 features)
    \item \textbf{NBR}: First and second harmonic amplitudes, first harmonic phase (cos), offset, and residual variance (5 features)
    \item \textbf{CRSWIR}: First and second harmonic phase (cos), offset, and residual variance (4 features)
\end{itemize}

Note that some features from the complete 32-feature set (4 indices × 8 metrics) were eliminated during recursive feature selection, resulting in this optimized 14-feature subset.

The inclusion of NBR's second harmonic amplitude proved critical—it captures the asymmetric green-up and senescence patterns characteristic of deciduous forests. Phase features encode peak timing as circular variables using cosine and sine components.

Our Random Forest classifier uses 50 trees with regularization parameters (max\_depth=30, min\_samples\_split=30) optimized through HalvingGridSearchCV—a successive halving approach that efficiently explores hyperparameter space by progressively eliminating poorly performing parameter combinations using increasing fractions of the training data. We selected Random Forest as our classifier based on its proven effectiveness in remote sensing applications, particularly for forest mapping where it consistently delivers high accuracy while avoiding overfitting to high-dimensional data (Belgiu \& Drăguţ, 2016). Cross-validation employs eco-region balanced folds, ensuring each ecological gradient contributes proportionally to model training and preventing overfitting to dominant regions.

\subsection{From Local Training to National Mapping}

The true test of our approach came in scaling from 14 million training pixels to France's 2+ billion forest pixels. Our processing pipeline parallelizes beautifully: each 100×100\,km tile runs independently, with 1024×1024 pixel chunks processed simultaneously across 80 CPU cores.

The entire workflow—from raw Sentinel-2 imagery to final probability maps—completes in 48 minutes for harmonic feature extraction plus under an hour for Random Forest inference. This efficiency enables annual updates, tracking forest changes as they happen rather than waiting years for new reference data.

\subsection{Validation Strategy}

To ensure our map's reliability, we implemented two validation approaches. First, spatial cross-validation using 2.5\,km tiles with at least 1\,km separation between training and test sets prevents overfitting to local patterns. Second, we compared our results against existing products (BD Forêt V2 and Copernicus DLT) using 10×10\,km blocks, computing agreement metrics only for pixels classified as forest in at least one product. This block-wise assessment reveals regional performance patterns and highlights where our harmonic approach excels—particularly in mixed deciduous-evergreen stands where spectral confusion is highest.

\section{Results}

\subsection{Finding the Goldilocks Zone: Why Two Harmonics?}

The first challenge in harmonic analysis is choosing the right level of complexity. Too few harmonics miss important phenological details; too many chase noise rather than signal. Our systematic comparison across France revealed that two harmonics hit the sweet spot (Figure \ref{fig:harmonic_cmp}).

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/harmonic_comparison.png}
    \caption{The evolution from simple to complex: one harmonic captures basic seasonality, two harmonics reveal asymmetric patterns typical of deciduous forests, while three harmonics begin fitting noise rather than phenology.}
    \label{fig:harmonic_cmp}
\end{figure}

Our comprehensive analysis examined one, two, and three harmonic configurations across all French eco-regions, evaluating signal-to-noise ratios, reconstruction quality, and energy distribution. The results reveal distinct patterns between forest phenology types and confirm the optimal configuration for national-scale mapping.

A single harmonic captures the basic annual cycle but forces symmetric spring green-up and autumn senescence—unrealistic for most deciduous species. Adding a second harmonic allows the model to represent rapid spring flush followed by gradual autumn decline, matching biological reality. The third harmonic, however, provides diminishing returns: our signal-to-noise analysis shows it primarily fits atmospheric contamination and registration artifacts rather than genuine phenological patterns (Figure \ref{fig:snr_maps}).

\subsubsection{Performance Metrics by Harmonic Configuration}

Our analysis of harmonic configurations reveals significant differences in performance between deciduous and evergreen forests. Table \ref{tab:harmonic_performance} presents the comprehensive comparison across configurations, showing how energy distribution and signal quality vary with phenology type.

\begin{table}[H]
    \centering
    \begin{tabular}{llccccc}
        \hline
        Configuration & Phenology & $SNR_{dB}$ (Median) & $SNR_{dB}$ (IQR) & $E_1$ & $E_2$ & $E_3$ \\
        \hline
        \multirow{2}{*}{1 Harmonic} 
            & Deciduous & 8.74 & \textbf{5.84} & 100.00 & -- & -- \\
            & Evergreen & 1.30 & \textbf{8.77} & 100.00 & -- & -- \\
        \hline
        \multirow{2}{*}{2 Harmonics} 
            & Deciduous & 10.23 & 5.94 & 91.14 & 8.86 & -- \\
            & Evergreen & 4.20 & 9.52 & 76.11 & 23.89 & -- \\
        \hline
        \multirow{2}{*}{3 Harmonics} 
            & Deciduous & \textbf{13.34} & 8.56 & 69.45 & 15.72 & 14.83 \\
            & Evergreen & \textbf{6.06} & 10.77 & 61.61 & 25.34 & 13.06 \\
        \hline
        \multicolumn{7}{l}{\small $E_1$, $E_2$, $E_3$: Energy percentage in first, second, and third harmonics} \\
        \multicolumn{7}{l}{\small SNR: Signal-to-noise ratio in decibels; IQR: Interquartile range} \\
    \end{tabular}
    \caption{Performance metrics for harmonic configurations by phenology type. Bold values indicate best performance for each metric.}
    \label{tab:harmonic_performance}
\end{table}

The breakdown by phenology reveals critical insights. Evergreen pixels consistently show lower SNR values and higher variability (larger IQR) compared to deciduous pixels, indicating that our harmonic model more effectively captures deciduous temporal signatures. This difference stems from the subtle seasonal variations in evergreen vegetation, which are harder to distinguish from noise in satellite observations.

Energy distribution analysis shows that deciduous forests concentrate more energy in the first harmonic (91.14\% vs 76.11\% for evergreen with two harmonics), reflecting their pronounced seasonal cycle. Evergreen forests require more energy in higher harmonics to capture their subtle variations, with the second harmonic accounting for 23.89\% of the signal energy compared to only 8.86\% in deciduous forests.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Frontiers LaTeX template – v2 (2025-04-10)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[utf8]{FrontiersinHarvard}
\usepackage{url,hyperref,lineno,microtype,subcaption}
\usepackage{natbib}
\usepackage[onehalfspacing]{setspace}
\usepackage{float}
\usepackage{multirow}
\linenumbers

\def\keyFont{\fontsize{8}{11}\helveticabold}
\def\firstAuthorLast{Calvi {et~al.}}
\def\Authors{Arthur Calvi\,$^{1,*}$, Sarah Brood\,$^{2}$, Co-Author\,$^{1,2}$, OpenAI Codex\,$^{3}$, Anthropic Claude Code\,$^{4}$}
\def\Address{$^{1}$ Laboratory X, Institute X, Department X, City X, Country X\\
$^{2}$ Laboratory Y, Institute Y, Department Y, City Y, Country Y\\
$^{3}$ OpenAI, San Francisco, CA, USA\\
$^{4}$ Anthropic, San Francisco, CA, USA}
\def\corrAuthor{Arthur Calvi}
\def\corrEmail{email@uni.edu}

@@

\title[France tree-phenology map]{France-wide 10\,m Deciduous–Evergreen Mapping: AlphaEarth Embeddings vs Harmonics}
%je mentionnerai le capteur dans le titre Mapping accuracy for evergreen and deciduous forests using AlphaEarth embeddings vs. sentinel 2 time series. ou Comparing model performances with AlphaEarth embeddings vs. sentinel 2 time series for classifying evergreen and decduous forests pas besoin de mettre ‘in France’

@@

\begin{abstract}
We present an operational route to annua (#a model to map … ), 10\,m deciduous–evergreen mapping for France. We compare two feature families under identical data and evaluation: (i) AlphaEarth foundation‑model embeddings (Emb‑Top14: 14 dimensions selected within folds from the 64‑band annual vectors) and (ii) (#features extracted from time series of Sentinel 2 images fitted with harmonic curves) physics‑informed harmonic descriptors (Harm‑Top14: 14 interpretable indices summarizing seasonal amplitude, timing, offset, and residuals). Using eco‑region–balanced (#rephrase by explaining what is an ecoregion, can be «  the features and weights have been balanced over different forest regions « ) folds and weights across 14.1M forest pixels, embeddings deliver modest, consistent gains (#gains of accuracy + cite the metrics)over harmonics (92.4\% vs 90.4\% OA) together with smoother, more coherent maps. Calibration is strong for both (ECE 0.033 vs 0.059), enabling stable thresholding (#je ne comprends pas cette phrase, la reécrre stp). Agreement with legacy products (DLT, BD Forêt) (# full names and refs) is comparable nationally while revealing expected regional divergences where annual phenology departs from static compilations. Because embeddings are pre‑computed and accessible in Earth Engine, labels—not features—are the practical bottleneck, making a country‑scale, annually updated phenology layer (#it is map of evergreen / deciduous forest types) achievable for downstream disturbance monitoring (#forest type is not necessarily for disturbance monitoring, can remove this part).

\keyFont{\section{Keywords:} deciduous-evergreen, AlphaEarth embeddings, harmonic analysis, Random Forest, phenology}
\end{abstract}

@@

% Teaser figure pinned within Introduction
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{images/sat_emb_classes.png}
    \caption{South‑east of France: (left) true‑color satellite view; (middle) learned embedding visualization with AlphaEarth dimensions A46, A18, A05 mapped to RGB; (right) classification probabilities where more orange indicates deciduous and more blue indicates evergreen. This teaser illustrates how embeddings accentuate phenological structure that drives the downstream classifier.}
    \label{fig:teaser_sat_emb_class}
\end{figure}

An annually refreshed deciduous–evergreen layer (#forest classification) at 10\,m would help stabilize forest disturbance detection (#be more general, help to monitor forest changes, in particular after disturbances when a forest type can change). Existing continental products illustrate the gap: Copernicus DLT classifies broadleaf versus conifer but cannot separate deciduous from evergreen broadleaf \citep{EU2024a} (#here explain in very few words what they use eg sentinel satellites); BD Forêt V2 offers detailed classes yet remains static (2007–2018) \citep{IGN2024}. Disturbance algorithms (CCDC, BEAST, BFAST, LandTrendr) are sensitive to phenological swings (#periodicity and abrupt changes) \citep{Zhu2014,Zhao2019,Verbesselt2010a,Verbesselt2010b,Kennedy2010,Kennedy2018} (#accurate ?); a current phenology prior would let analysts tighten thresholds for deciduous pixels (high natural variance) and relax them for evergreen stands (low variance), reducing false alarms while highlighting genuine change.

Foundation models provide a practical way to build such priors. Recent efforts (SatMAE, Prithvi-EO, AlphaEarth) train on multi-temporal satellite imagery and transfer well to downstream tasks \citep{Cong2022,Szwarcman2024PrithviEO2,AlphaEarth2025} (#explain main satellite and other data used). AlphaEarth, in particular, publishes annual 64-dimensional embeddings at 10\,m. Because these vectors are precomputed and temporally consistent, practitioners can treat them as a ready-made feature stack; the learned spatial context makes them well suited to simple per-pixel classifiers, shifting the bottleneck from feature engineering to assembling reliable labels.

Against this backdrop we compare two compact feature sets under identical data and evaluation: Emb-Top14 (fourteen dimensions selected within cross-validation folds from the 64D AlphaEarth vectors) and Harm-Top14 (fourteen (#how many bands or indices from the satellite) physics-informed descriptors summarizing seasonal amplitude, timing, offset, and residual variance of Sentinel-2 indices). We ask whether embeddings can compete with carefully engineered harmonics while simplifying the pipeline, whether embedding maps remain smoother without additional filtering (#why is smoother derirable for a good classification ?), and whether the resulting confidence scores are stable enough for practical thresholding (#last part is jargon, please rewrite. Better have one sentence per question).

Our processing chain is intentionally simple: tile mainland France into 2.5\,km cells; assemble 14.1M labeled forest pixels with eco-region tags and area-aware weights; compute Emb-Top14 or Harm-Top14 features; train identical Random Forests with eco-region–stratified (#was not explained), tile-grouped cross-validation; and evaluate accuracy, spatial coherence, and confidence, including comparisons to DLT and BD Forêt restricted to forest pixels. (#rewrite and break the sentence into a classical structure. The model chosen for classification, data preparation, training and cross validation + then a new sentence for the comparison with other products)

Our contributions are threefold: (i) a France-wide, annual 10\,m phenology map (#in the whole paper replace « phenology’ » by evergreen-deciduous classification ( you can use PC for phenology classification in case the same word is used many times) trained and evaluated with eco-region–balanced folds and weights (#unclear to readers); (ii) a controlled comparison of Emb-Top14 and Harm-Top14 under identical data and Random Forests (#with the same train and test data), including spatial-coherence and confidence analyses; and (iii) a national comparison with DLT and BD Forêt, with eco-region breakdowns that explain where annual phenology diverges from static products (#our annual PC maps).

Embedding-based approaches have already shown competitive performance across EO tasks (#applications) \citep{Cong2022,Xie2024FoundationEffective}, and pre-trained representations are increasingly used for rapid environmental monitoring (#can you cite here some concrete aplications)\citep{Szwarcman2024PrithviEO2}. Dataset design also matters: phenology-informed, globally uniform pretraining improves ecological transfer (#what does the last word means ?) \citep{Plekhanova2025SSL4Eco}. Nevertheless, questions remain about interpretability and cross-regional reliability \citep{Xie2024FoundationEffective}. By benchmarking embeddings against explicit harmonics under matched conditions (#with the same clasification model set up), we quantify the trade-offs between accuracy, interpretability, and operational feasibility (#not sure this is done in the paper, but see …

quantiying interpretabilty is diffrent from comparing two models ).

Recent studies reinforce the motivation. Houriez et al. extend LANDFIRE (#is it a land cover ? ) vegetation types into Canada with simple classifiers on AlphaEarth features \citep{Houriez2025AEFDataGen}; a Siamese U-Net on AlphaEarth imagery reports strong burned-area (#was used for mapig burned area ? ) mapping across continents \citep{Seydi2025AlphaEarthBurnedArea}; and national phenology mapping efforts (#is it land cover maps or phenology ?) continue to emphasize temporal descriptors for evergreen discrimination \citep{Inglada2017,Li2023,Low2020,Bolton2020}. Our comparison situates harmonic analysis and AlphaEarth embeddings within this landscape: harmonics provide interpretable seasonal signals useful for scientific inquiry, while embeddings offer scalable, precomputed features that agencies can deploy immediately.

@@

\subsection{Dataset composition and sampling}
\section{Methods}

\subsection{Study area and eco-regional stratification}
\label{sec:greco}
We focus on mainland France and Corsica, where temperate, montane, and Mediterranean forest types co-exist within short distances. To preserve ecological diversity in both training and evaluation we rely on the eleven \emph{Grandes Régions Écologiques} (GRECO) defined by the French National Forest Inventory \citep{IGN2013GRECO}. These eco-regions capture the dominant gradients from Atlantic mixed forests through semi-continental plateaus to Mediterranean evergreen stands and Alpine massifs. We partitioned the national forest mask into 639 non-overlapping 2.5\,km\,$\times$\,2.5\,km tiles (Figure~\ref{fig:training_tiles}), retaining tiles that contain at least ten labelled pixels and are separated by 5\,km to limit spatial autocorrelation. Sample weights compensate for the unequal forest area of each eco-region: for region \(r\), the weight \(w_r\) equals the ratio between its effective forest area fraction \(A_r/\sum_{j} A_j\) and the observed sample fraction \(n_r/N\). We normalise the weights so that \(\sum_i w_i = N\); the resulting weights range from 0.89 in densely sampled regions to 1.48 in rare mountainous tiles.

The GRECO framework offers concise descriptors that explain the spatial behaviour of our models. Region A (Grand Ouest cristallin et océanique) is a humid Atlantic bocage of oak–chestnut forests interspersed with post-war conifer plantations and dense hedgerows. Region B (Centre Nord semi-océanique) covers the Paris Basin cuestas with extensive pedunculate/sessile oak and hornbeam on loess plateaus and Scots pine on sandy soils. Region C (Grand Est semi-continental) spans the Ardennes and Lorraine, with beech–fir forests on cool uplands. Region D (Vosges) combines silver fir–beech on steep crystalline slopes, while Region E (Jura) hosts calcareous beech–fir mosaics above karst plateaus. Region F (Sud-Ouest océanique) includes the Landes maritime pine massif and humid mixed stands. Region G (Massif Central) contains montane beech–fir and chestnut belts, Region H (Alps) sharply transitions from deciduous foothills to coniferous subalpine belts, and Region I (Pyrenees) alternates Atlantic and Mediterranean influences over steep gradients. Region J (Mediterranean) is dominated by evergreen holm and cork oak maquis with summer drought stress, and Region K (Corsica) blends lowland maquis with high-elevation Laricio pine forests. These descriptors provide the ecological backdrop for the regional analyses reported in Section~\ref{sec:results} and are summarised in Table~\ref{tab:greco_summary}.

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{p{0.9cm}p{3.2cm}p{4.1cm}p{5.2cm}}
        \toprule
        \textbf{Code} & \textbf{Region} & \textbf{Climate and terrain} & \textbf{Dominant forest structure} \\
        \midrule
        A & Grand Ouest cristallin et océanique & Humid Atlantic plains and low hills, dense hedgerow landscapes & Oak–chestnut coppice with maritime pine and Sitka spruce plantations \\
        B & Centre Nord semi-océanique & Loess plateaus and chalk cuestas of the Paris Basin & Pedunculate/sessile oak with hornbeam/beech; Scots and Corsican pine on sandy soils \\
        C & Grand Est semi-continental & Ardennes and Lorraine uplands with colder winters & Beech–fir and spruce forests on mesic plateaus, mixed oak lowlands \\
        D & Vosges & Steep crystalline range with high precipitation & Silver fir–beech high forests with spruce and Douglas-fir plantings \\
        E & Jura & Limestone plateaus, cool montane climate & Calcareous beech–fir mosaics and mixed spruce on karst slopes \\
        F & Sud-Ouest océanique & Landes coastal plain and Atlantic piedmont & Maritime pine estates interleaved with humid oak, alder, and chestnut stands \\
        G & Massif Central & Volcanic plateaus and valleys with montane climate & Beech–fir belts, chestnut groves, Scots and Douglas pine on poorer soils \\
        H & Alps & Strong elevational gradient, deep glacial valleys & Deciduous foothills (oak, beech) grading to spruce–larch–fir subalpine belts \\
        I & Pyrenees & Atlantic–Mediterranean transition, steep valleys & Oak/beech foothills with fir-spruce upper slopes and Mediterranean pine on southern flanks \\
        J & Mediterranean & Coastal ranges and plateaus with summer drought & Evergreen holm and cork oak maquis, Aleppo and maritime pine stands \\
        K & Corsica & Crystalline massif with rugged relief & Lowland evergreen maquis and extensive Laricio pine forests above 900\,m \\
        \bottomrule
    \end{tabular}
    \caption{Summary of the eleven GRECO eco-regions used for sampling and evaluation.}
    \label{tab:greco_summary}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{images/tiles_2_5_km_final_visualization.png}
    \caption{Distribution of the 639 eco-region balanced training tiles. Warm colours indicate Mediterranean and montane eco-regions, while cool colours denote Atlantic and semi-continental domains. Each tile contains 10\,m pixels labelled as deciduous or evergreen and is used as an indivisible unit during cross-validation.}
    \label{fig:training_tiles}
\end{figure}

\subsection{Reference labels}
The supervised dataset aggregates 14{,}086{,}937 forest pixels (10\,m resolution) drawn from four complementary sources. (i) PureForest provides 135\,000 lidar-guided patches with expert-verified dominant species, mainly covering monospecific stands in southern France \citep{gaydon2024pureforestlargescaleaeriallidar}. (ii) The RENECOFOR long-term monitoring network supplies plot inventories with tree measurements above 15\,m height collected between 2019 and 2020 \citep{ulrich:hal-03444393}. (iii) The Tree Position Calibration campaign geolocates dominant trees with airborne lidar to refine field GPS positions and species attribution \citep{ONF,IGN_LiDARHD}. (iv) BD Forêt V2 contributes mapped stands ( > 5{,}000\,m$^2$) updated between 2005 and 2019 \citep{IGN2024}. We apply a 100\,m negative buffer to BD Forêt polygons to mitigate edge drift before rasterisation. Each pixel inherits its eco-region and tile identifier; categorical attributes (phenology, genus, species, source, acquisition year) are encoded via consistent look-up tables. Deciduous pixels represent 75.5\,\% of the samples (10{,}639{,}124 pixels), evergreens 24.5\,\% (3{,}447{,}813 pixels). Buffered BD Forêt polygons supply 88.6\,\% of the pixels, while in-situ inventories anchor 11.4\,\%, ensuring we retain authoritative field labels within every eco-region.

\subsection{Feature extraction}
\subsubsection{Sentinel-2 harmonic descriptors}
We reconstruct Sentinel-2 Level-2A surface reflectances from Google Earth Engine monthly MEDIAN composites, applying the s2cloudless cloud-probability mask with a 75\,\% threshold, removing QA60 bits 10 and 11 (opaque and cirrus clouds), and discarding scenes with more than 95\,\% cloudy pixels. No gap filling or scene classification layer is applied. For each tile we derive the vegetation indices NDVI, EVI, NBR, and the SWIR ratio (CRSWIR). The annual signal of each index is fitted with two sinusoidal harmonics through ordinary least squares,
\[
  x(t) = C + \sum_{k=1}^{2} \big[a_k \cos\!\big(2\pi k t/T\big) + b_k \sin\!\big(2\pi k t/T\big)\big], \quad T = 1\,\text{year}.
\]
We recover offsets, amplitudes, phases, and residual variance; phases are transformed into their sine and cosine components to avoid wrap-around discontinuities. Recursive feature elimitation (Section~\ref{subsubsec:rfe}) retained 14 descriptors (Table~\ref{tab:harmonic14}) that balance interpretability and cross-validated accuracy (Supplementary Section~S1).

\begin{table}[H]
    \centering
    \small
    \begin{tabular}{lp{0.62\textwidth}}
        \hline
        \textbf{Feature} & \textbf{Ecological signal} \\
        \hline
        NDVI first-harmonic amplitude & Seasonal strength of broadleaf greenness pulses \\
        NDVI first-harmonic phase (cosine/sine) & Timing of green-up and senescence transitions \\
        NDVI second-harmonic phase (sine) & Asymmetry between rapid spring and gradual autumn trajectories \\
        NDVI offset & Mean canopy greenness throughout the year \\
        NBR first-harmonic amplitude & Annual moisture and structural contrast between canopies and soil \\
        NBR first-harmonic phase (cosine) & Calendar timing of minimum fuel moisture \\
        NBR second-harmonic phase (cosine) & Secondary moisture cycle in bimodal climates \\
        NBR offset & Average woody biomass signal \\
        NBR residual variance & Short-term disturbance departures from harmonic behaviour \\
        CRSWIR first-harmonic phase (cosine) & Timing of peak shortwave-infrared water stress \\
        CRSWIR second-harmonic phase (cosine) & Recovery pattern of evergreen water content \\
        CRSWIR offset & Mean canopy water and lignin content \\
        CRSWIR residual variance & Fine-scale heterogeneity in SWIR response \\
        \hline
    \end{tabular}
    \caption{Harmonic descriptors retained for the HARM-14 feature set. Phases are represented through sine and cosine components to avoid angular discontinuities.}
    \label{tab:harmonic14}
\end{table}

\subsubsection{AlphaEarth embeddings}
AlphaEarth provides annual 64-dimensional embeddings at 10\,m resolution, learned from multi-sensor stacks (Sentinel-1/2 optical reflectances, Sentinel-1 backscatter, Landsat seasonal composites, GEDI structure, ERA5-Land climate predictors) by masking and reconstructing spatio-temporal tokens \citep{AlphaEarth2025}. The resulting vector can be viewed as a “spectro-temporal fingerprint” of each pixel: it summarises how the site looks across the year, how it relates to its neighbourhood, and how ancillary drivers (elevation, climate) modulate that pattern. Because the embeddings are already computed and distributed through Earth Engine, practitioners can treat them as off-the-shelf features without training deep networks. We retrieved the 2018, 2020, 2022, and 2023 embedding rasters, converted the tiles to Parquet, and removed pixels with non-finite components (mostly coastline artifacts). To align with the harmonic dataset we perform an inner join on tile, row, and column indices so that both feature families use the exact same labelled pixels and eco-region weights. The 14 embedding dimensions retained after feature elimination (Section~\ref{subsubsec:rfe}) define the \textbf{EMB-14} feature set used in training and evaluation.
